Parsing Postfix log files

### 2008-10-29 doc/logparser-implementation.tex
Second half of the corrections to section 5.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
due to forwarding, or internally generated by Postfix.  Local submission is
the simplest case: a queueid is assigned immediately and the sender address
is logged (action: PICKUP\@; flowchart:~2).

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in \sectionref{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections, so discussion
is deferred to \sectionref{tracking re-injected mail}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
\sectionref{identifying-bounce-notifications} for details).  Bounce
notifications are the primary example of internally generated mails, though
there may be other types.\footnote{Postfix may generate mails to the
administrator when it encounters configuration errors, but such mails are
presumably rare.}

@@

    \item Postfix will log the sender and recipient addresses separately
        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).

@@

\subsubsection{Tracking re-injected mail}

@@

    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
        bounce messages.  This action creates a new mail if necessary; if
        the mail already exists the unknown origin flag will be removed.
        The action also marks the mail as a bounce notification.  To deal
        with complication \sectionref{Further out of order log lines} this
        action checks a cache of recent bounce mails to avoid creating
        bogus bounce mails when lines are logged out of order.

    \item [COMMIT] Enter the data from the mail into the database. Entry
        will be postponed if the mail is a child waiting to be tracked.
        Once entered, the mail will be deleted from the state tables.
        Deletion will be postponed if the mail is the parent of re-injected
        mail.

    \item [MAIL\_QUEUED] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
        out of order log lines; see \sectionref{discarding cleanup lines}
        for details.

    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
        than the local server accepts, the mail will be discarded, and the
        client informed.  See TIMEOUT for further discussion; the two are
        handled in exactly the same way.

Adding new actions is not as simple as adding new rules, though care has
been taken in the parser's design and implementation to make adding new
actions as painless as possible.  The implementor writes a subroutine that
accepts the standard arguments given to actions, and registers it as an
action\footnote{The new action must be registered before the rules are
loaded, as it is an error for a rule to reference an unregistered action;
this helps catch mistakes made when adding new rules.} by calling the
framework subroutine add\_actions() with the name of the new action
subroutine as a parameter.  No other work is required on the implementor's
part to integrate the action into the parser; all of their attention and
effort can be focused on the correctness of their action.  The only
negative aspect is that the process involves editing the parser source
code, which makes upgrading to a later version of the parser more
difficult, though by no means impossible.  If the author of the new action
wishes, they can take advantage of the parser's object oriented
implementation~\cite{Wikipedia-object-orientation} by subclassing it and
implementing their changes in the derived class, allowing future upgrading
of the parser with greatly reduced chance of conflicts.\footnote{The real
difference between the two approaches is where the new code is placed.  The
simpler option is to change the parser code directly, but those changes
will then have to be made to subsequent versions of the parser, and as the
scope of the changes increases so does the chance of conflict, or mistakes
when copying the action.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
addition creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the log files used
exhibit the idiosyncrasies and peculiarities a mail server in the wild must
deal with, but the downside is that significantly altering the
configuration just to log messages from a different Postfix component is
not an option.}

\subsubsection{Summary}

    \item Two state tables to hold the data structure for each connection.

Multiple independent mails may be delivered during one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log entries will not be correlated by \pid{} any
more (its queueid will be used instead), so there is no longer any
ambiguity about which mail a given log line belongs to (except when
timeouts occur during the data phase
\sectionref{timeouts-during-data-phase}).  If the original connection has
any interesting data (e.g.\ rejections) it will be saved to the database
when the client disconnects.  One unsolved difficulty is distinguishing
between different groups of rejections, e.g.\ when dealing with the
following sequence:

    \item The client issues the RSET command to reset the session.

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this occurs
a child mail will be injected into the Postfix queue, but without the
explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
Thus the source is not immediately discernible from the log line in which
the mail first appears; from a strictly chronological reading of the log
files it usually appears as if the child mail has appeared from thin air.
Subsequently the parent mail will log the creation of the child mail:

The child mail has been created with queueid \texttt{56F5B43FD}.  Different
delivery methods result in different log lines:

    \item [Local delivery:] delivered to command:
        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation log line!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered by the parent mail once it has been
identified.

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail that lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in \sectionref{tracking re-injected
mail}, but unlike the solution described therein bounce notifications do
not eventually have a log line that identifies their source.  Heuristics
must be used to identify bounce notifications, and those heuristics are:

@@

    \item The sender address is $<>$.

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \SMTP{} or submitted
        locally by \daemon{postdrop} or sendmail.

    \item The message-id has a specific format: \newline{}
        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing.  In the latter case the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format (e.g.\ if it was generated by
        Postfix on this or another server) the queueid in the message-id
        will not match the queueid of the mail.

@@

Once a mail has been identified as a bounce notification, the unknown
origin flag is cleared and the mail can be cleaned up and entered in the
database.

@@

    \item The sender address must be $<>$.

@@

If a mail is mis-classified as a bounce message it will almost certainly
have been generated internally by Postfix; arguably mis-classification in
this case is a benefit rather than a drawback, as other mails generated
internally by Postfix will be handled correctly.

Postfix 2.3 (and hopefully subsequent versions) log the creation of a
bounce message.

@@

Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.\footnote{Microsoft Outlook is
one client that behaves in this fashion; no attempt has been made to
identify other clients that act in a similar way.}  An example dialogue
exhibiting this behaviour is presented below (lines starting with a three
digit number are sent by the server, the other lines are sent by the
client):

@@

    \item Are there two or more \daemon{smtpd} log lines?  Does the second
        result have a postfix\_action attribute of ACCEPTED\@?  The first
        two \daemon{smtpd} log lines will be a connection log line and a
        mail acceptance log line (Postfix logs acceptance as soon as the
        first recipient has been accepted).

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

These checks are made during the DISCONNECT action: if all checks are
successful then the mail is assumed to have been discarded when the client
disconnected.  There will be no further entries logged for such mails, so
without identifying and entering them in the database immediately they
accumulate in the state table and will cause clashes if the queueid is
reused.

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake \textit{timeout\/} will be used
throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the data to the database, but if a
timeout occurs there is no data available to save; the timeout is recorded
with the connection data instead.

@@

A timeout may thus apply either to an accepted mail or a rejected mail.  To
distinguish between the two cases the algorithm compares the timestamp of
the last accepted mail against the timestamp of the last line logged by
\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
CLONE action keeping a list of all mails accepted on each connection).  If
the mail acceptance timestamp is later then the timeout applies to the
just-accepted mail, which will be discarded.  If the \daemon{smtpd}
timestamp is later there was a rejection between the accepted mail and the
timeout: the action assumes that the timeout applies to a rejected delivery
attempt and finishes.  This assumption is not necessarily correct, because
Postfix may have accepted an earlier recipient and rejected a later one, in
which case the timeout applies to the accepted mail, which should be
discarded.  This has not been a problem in practice, though it may be in
future.

This complication is further complicated by the presence of out of order
\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
details.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} line is logged after the timeout line; parsing this line
causes the creation of a new state table entry for the queueid in the log
line.  This is incorrect because the line actually belongs to the mail
that has just been discarded; the next log line for that queueid will be
seen when the queueid is reused for a different mail, causing a queueid
clash and the appropriate warning.

In the case where the \daemon{cleanup} line is still pending during the
TIMEOUT action the algorithm updates a global cache of queueids, adding the
queueid and the timestamp from the timeout line.  When the next
\daemon{cleanup} line is parsed for that queueid the cache will be checked
(during the MAIL\_QUEUED action),
and the line will be deemed part of the discarded mail and discarded if it
meets the following requirements:

@@

    \item The queueid must not have been reused yet, i.e.\ there is not an
        entry in the state tables for the queueid.

    \item The timestamp of the \daemon{cleanup} line must be within ten
        minutes of the mail acceptance timestamp.  Timeouts happen after
        five minutes, but some data may have been transferred slowly
        (perhaps because either the client or server is suffering from
        network congestion or rate limiting), and empirical evidence shows
        that ten minutes is not unreasonable; hopefully it is a good
        compromise between false positives and false negatives.

@@

The next \daemon{cleanup} line must meet the criteria above for it to be
discarded because not every connection where a timeout occurs will have a
\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
next \daemon{cleanup} line after a TIMEOUT it would in some cases be
mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
queueid will be removed from the cache of timeout queueids when the next
\daemon{pickup} line containing that queueid is parsed.

@@

Occasionally the \daemon{pickup} line logged when mail is submitted locally
via sendmail appears later in the log file than the \daemon{cleanup} line
for that mail.  This seems to occur during periods of particularly heavy
load, so is most likely due to process scheduling vagaries.  Normally if
the queueid given in the PICKUP line exists in the state tables a warning
is generated by the \daemon{pickup} action, but if the following conditions
are met it is assumed that the lines were out of order:

@@

As always with heuristics there may be circumstances in which these
heuristics match incorrectly,  but none have been identified so far.

@@

Occasionally a \daemon{smtpd} will just stop logging, without an
immediately obvious reason.  After poring over log files for some time
there are several reasons for this infrequent occurrence:

@@

    \item Postfix is stopped or its configuration is reloaded.  When this
        happens all \daemon{smtpd} processes exit, and all entries in the
        connections state table must be cleaned up, entered in the database
        if there is sufficient data, and deleted.  The queueid state table
        is untouched.

@@

@@

The above descriptions appear to cover all situations identified thus far
where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
active connection the last accepted mail may need to be discarded, as
detailed in \sectionref{timeouts-during-data-phase}.

These occurrences are handled by the three actions POSTFIX\_RELOAD,
SMTPD\_DIED and SMTPD\_WATCHDOG\@.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
occurs only five times in 60,721,709 log lines, but for completeness of the
algorithm it should be dealt with.  The five occurrences in the test log
files have the same characteristics: the \daemon{local} log line showing
delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
removal of the mail from the queue because delivery is completed.  This
causes problems: the mail is not complete, so entry into the database
fails; a new mail is created when the \daemon{local} line is parsed and
remains in the state tables; four warnings are issued per pair of out of
order log lines.

The solution to this problem is to examine the list of programs that have
logged lines for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The
SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it is
found; if the additional log lines have caused the mail to reach a valid
combination entry will proceed, otherwise it must be postponed once more.

@@

@@

    \item [\daemon{local}:] Local delivery of a bounce notification, or
        local delivery of a forwarded or tracked mail.

@@

    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
        submitted mail, \newline{} both local and remote delivery.

@@

    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
        delivery only.

    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
        a bounce notification.

    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
        client, then remotely delivered (typically relaying mail for
        clients on the local network).

@@

This check applies to accepted mails only, not to rejected mails.

@@

    \item The timestamp of the last log line for the mail must be 12 hours
        or more earlier than the last log line parsed.

    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
        log entries for the mail, with no additional log entries.

@@

12 hours is a somewhat arbitrary time period, but it is far longer than
Postfix would delay delivery of a mail in the queue.\footnote{This may be a
problem if Postfix is not running for an extended period of time.}  The
state tables are scanned for mails matching the characteristics above each
time the end of a log file is reached, and matching mails are deleted.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally
mails in the process of being delivered will be deleted.  This results in
the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
removed from the state tables and saved in the database.  The DELETE action
adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
action, and the current log line discarded if the following conditions are
met:

@@

    \item The timestamp of the log line is within 5 minutes of the end time
        of the mail.

>>>>

XXX SHOULD I SAY ``cleaned up'' EVERYWHERE OR NOT\@?

@@

XXX SHOULD THESE BE SEPARATE SUBSUBSECTIONS\@?

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \SMTP{}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own; it is explained in \sectionref{Re-injected mails}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics as described in
\sectionref{identifying-bounce-notifications}.  Bounce notifications are
the primary example of internally generated mails, though there may be
other types.  Postfix may generate mails to the administrator when it
encounters configuration errors, but such mails are presumably rare.

@@

    \item Postfix will log the sender and recipient addresses separately
        (action: SAVE\_DATA\@; flowchart:~9).

@@

XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
UPDATED, AT WORST IT'S RUBBISH\@.

\subsection{Actions}

\label{actions in implementation}

XXX INTRODUCTION\@.

%\begin{eqlist}[\def\makelabel#1{\bfseries#1:}]

XXX IMPROVE HOW THIS LOOKS

All data extracted by the rule will be automatically saved, regardless of
the action which is invoked.

XXX MAKE ALL ACTION DESCRIPTIONS FLOW BETTER\@.

    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
        bounce messages.  This action creates a new mail if necessary; if
        the mail already exists the unknown origin flag will be removed.
        The action also marks the mail as a bounce notification.  To deal
        with complication \sectionref{Bounce notification mails delivered
        before their creation is logged} this action checks a cache of
        recent bounce mails to avoid creating bogus bounce mails when lines
        are logged out of order.

    \item [COMMIT] Enter the data from the mail into the database. Entry
        will be postponed if the mail is a child waiting to be tracked.
        Once entered, the mail will be deleted from the state tables.
        Deletion will be postponed if the mail is the parent of re-injected
        mail (\sectionref{Re-injected mails}).

    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
        the queue to deliver. This action is used for both \daemon{qmgr}
        and \daemon{cleanup} as it needs to deal with out of order log
        lines; see \sectionref{discarding cleanup lines} for details.  XXX
        EXPLAIN THIS BETTER\@.

    \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
        the local server accepts it will be discarded and the client
        informed of the problem.  This is handled in exactly the same way
        as the TIMEOUT action, further details are given there.

XXX HALF OF THIS CONTENT IS IN FOOTNOTES, WHICH SHOULD BE INLINED\@.

Adding new actions is not as easy as adding new rules, though care has been
taken in the parser's design and implementation to make adding new actions
as painless as possible.  The implementor writes a subroutine that accepts
the standard arguments given to actions, and registers it as an action by
calling the framework subroutine add\_actions() with the name of the new
action as a parameter.  The new action must be registered before the rules
are loaded, because it is an error for a rule to specify an unregistered
action; this helps catch mistakes made when adding new rules.  No other
work is required from the implementor to integrate the action into the
parser; all of their attention and effort can be focused on correctly
implementing their action.  The only negative aspect is that the process
involves editing the parser source code, which makes upgrading to a later
version of the parser more difficult, though by no means impossible.  If
the author of the new action wishes, they can take advantage of the
parser's object oriented implementation by subclassing it and implementing
their changes in the derived class, allowing future upgrades of the parser
with greatly reduced chance of conflicts.\footnote{The real difference
between the two approaches is where the new code is placed.  The simpler
option is to change the parser code directly, but those changes will then
have to be made to subsequent versions of the parser, and as the scope of
the changes increases so does the chance of conflict, or mistakes when
copying the action.  The more time consuming option is to write a subclass
containing the new actions and change the program that invokes the parser
so that it uses the subclass rather than the parser; the changes required
to the program invoking the parser are minor and much less likely to lead
to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser can
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
new action creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit of using this server is that
the log files used exhibit the idiosyncrasies and peculiarities a mail
server in the wild must deal with, but the downside is that significantly
altering the configuration just to log messages from a different Postfix
component is not an option.}

XXX IMPROVE THE INTRO\@: ``needs more of a lead-in''.

XXX SHOULD I HAVE A LIST OF HOW MANY COMPLICATIONS ARE HANDLED BY EACH
ACTION\@?

    \item Two state tables to hold the data structure for each connection
        and mail.

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log entries will not be correlated by \pid{} any
more, its queueid will be used instead (except when timeouts occur during
the data phase \sectionref{timeouts-during-data-phase}).  If the original
connection has any useful data (e.g.\ rejections) it will be saved to the
database when the client disconnects.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

    \item The client issues the RSET command to reset the \SMTP{} session.

\label{Re-injected mails}

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this
occurs a child mail will be injected into the Postfix queue, but without
the explicit logging that \daemon{smtpd} or \daemon{postdrop} injected
mails have.  Thus the source of the mail is not immediately discernible
from the log line in which the mail first appears; from a strictly
chronological reading of the log files it usually appears as if the child
mail has appeared from thin air.  Subsequently the parent mail will log the
creation of the child mail, e.g.\ parent mail 3FF7C4317 creates child mail
56F5B43FD\@:

\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
\tab{} delay=0, status=sent (forwarded as 56F5B43FD)}

XXX DO I NEED THIS LIST\@? IMPROVE HOW IT LOOKS IF I KEEP IT\@.

Different delivery methods result in different log lines:

    \item [Local delivery:] delivered to command: /usr/bin/procmail

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation log line!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered once the parent mail has been identified.

XXX MERGE THE REMAINDER OF THIS SECTION INTO THE PRIOR MATERIAL\@.

subsubsection{Tracking re-injected mail}

@@

@@

Postfix 2.2.x (and presumably previous versions) does not generate a log
line when it generates a bounce notification; suddenly there will be log
entries for a mail that lacks an obvious source.  There are similarities to
the problem of identifying re-injected mails discussed in
\sectionref{tracking re-injected mail}, but unlike the solution described
therein bounce notifications do not eventually have a log line that
identifies their source.  Heuristics must be used to identify bounce
notifications, and those heuristics are:

@@

    \item The sender address is \verb!<>!.

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \SMTP{} or submitted
        locally by \daemon{postdrop}.

    \item The message-id has a specific format: \newline{}
        \tab{} \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing (described in \sectionref{Re-injected mails}).
        In the latter case the message-id will be unchanged from the
        original bounce notification, and so even if it happens to be in
        the correct format (e.g.\ if it was generated by Postfix on this or
        another server) it will not match the queueid of the mail.

@@

Once a mail has been identified as a bounce notification, the unknown
origin flag is cleared and the mail can be entered in the database.

@@

    \item The sender address must be \verb!<>!.

@@

If a mail is mis-classified as a bounce message it will almost certainly
have been generated internally by Postfix; arguably mis-classification in
this case is a benefit rather than a drawback, as other mails generated
internally by Postfix will be handled correctly.  Postfix 2.3 (and
hopefully subsequent versions) log the creation of a bounce message.

@@

Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.  Microsoft Outlook is one
client that behaves in this fashion; other clients may act in a similar
way.  An example dialogue exhibiting this behaviour is presented below
(lines starting with a three digit number are sent by the server, the other
lines are sent by the client):

@@

    \item XXX IS THIS CORRECT NOW\@?  Are there two or more \daemon{smtpd}
        log lines?  Does the second result have a postfix\_action attribute
        of ACCEPTED\@?  The first two \daemon{smtpd} log lines will be a
        connection log line and a mail acceptance log line (Postfix logs
        acceptance as soon as the first recipient has been accepted).

@@

XXX MAYBE FIND OUT WHAT CLIENTS DISCONNECT\@?  Some mail clients disconnect
abruptly if a second or subsequent recipient is rejected; they may also
disconnect after other errors, but such disconnections are either
unimportant or are handled elsewhere in the parser
(\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not log a
message saying the mail has been discarded, as should be expected by now.
The checks to identify this happening are:

@@

If all checks are successful then the mail is assumed to have been
discarded when the client disconnected.  There will be no further entries
logged for such mails, so without identifying and entering them in the
database immediately they accumulate in the state table and will cause
clashes if the queueid is reused.

These checks are made during the DISCONNECT action.

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake \textit{timeout\/} will be used
throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded with the connection data instead, which is saved.

@@

XXX MAKE THIS PARAGRAPH CLEARER\@.  A timeout may thus apply either to an
accepted mail or a rejected mail.  To distinguish between the two cases the
algorithm compares the timestamp of the last accepted mail against the
timestamp of the last line logged by \daemon{smtpd} for that connection
(the TIMEOUT action is dependant on the CLONE action keeping a list of all
mails accepted on each connection).  If the mail acceptance timestamp is
later then the timeout applies to the just-accepted mail, which will be
discarded.  If the \daemon{smtpd} timestamp is later there was a rejection
between the accepted mail and the timeout: the action assumes that the
timeout applies to a rejected delivery attempt and finishes.  This
assumption is not necessarily correct, because Postfix may have accepted an
earlier recipient and rejected a later one, in which case the timeout
applies to the accepted mail, which should be discarded.  This has not been
a problem in practice, though it may be in future.  This complication is
further complicated by the presence of out of order \daemon{cleanup} lines:
see \sectionref{discarding cleanup lines} for details.

This complication is dealt with in the TIMEOUT action.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} line for a mail being accepted is logged after the timeout
line; parsing this line causes the MAIL\_QUEUED action to create a new
state table entry for the queueid in the log line.  This is incorrect
because the line actually belongs to the mail that has just been discarded;
the next log line for that queueid will be seen when the queueid is reused
for a different mail, causing a queueid clash and the appropriate warning.

When the \daemon{cleanup} line is still pending during the TIMEOUT action,
the action updates a global list of queueids, adding the queueid and the
timestamp from the log line.  When the next \daemon{cleanup} line is parsed
for that queueid the list will be checked (during the MAIL\_QUEUED action),
and the log line will be deemed part of the mail where the timeout occurred
and discarded if it meets the following requirements:

@@

    \item The queueid must not have been reused yet, i.e.\ it does not have
        an entry in the state tables.

    \item The timestamp of the \daemon{cleanup} log line must be within ten
        minutes of the mail acceptance timestamp.  Timeouts happen after
        five minutes, but some data may have been transferred slowly
        (perhaps because either the client or server is suffering from
        network congestion or rate limiting), and empirical evidence shows
        that ten minutes is not unreasonable; hopefully it is a good
        compromise between false positives (log lines incorrectly
        discarded) and false negatives (new state table entries incorrectly
        created).

@@

The next \daemon{cleanup} line must meet the criteria above for it to be
discarded because some, but not all connections where a timeout occurs will
have an associated \daemon{cleanup} line logged; if the algorithm blindly
discarded the next \daemon{cleanup} line after a timeout it would sometimes
be mistaken.  When the next \daemon{pickup} line containing that queueid is
parsed the queueid will be removed from the cache of timeout queueids,
regardless of whether it meets the criteria above.

This complication is handled by the TIMEOUT and MAIL\_QUEUED actions.

@@

When mail is submitted locally, \daemon{pickup} accepts the new mail and
generates a log line showing the source.  Occasionally this log line will
occur later in the log file than the \daemon{cleanup} log line, so the
PICKUP action will find that a state table entry exists for that queueid.
Normally if the queueid given in the PICKUP line exists in the state tables
a warning is generated by the \daemon{pickup} action, but if the following
conditions are met it is assumed that the lines were out of order:

@@

As always with heuristics there may be circumstances in which these
heuristics match incorrectly, but none have been identified so far.  This
complication seems to occur during periods of particularly heavy load, so
is most likely caused by process scheduling vagaries.

@@

Occasionally a \daemon{smtpd} will just stop logging, without an
immediately obvious reason.  After poring over log files for some time
several reasons have been found for this infrequent occurrence:

@@

    \item Postfix is stopped or its configuration is reloaded.  When this
        happens all \daemon{smtpd} processes exit, so all entries in the
        connections state table must be cleaned up, entered in the database
        if there is sufficient data, and deleted.

@@

    \item Every Postfix process uses a watchdog which kills the process if
        it is not reset for a considerable period of time (five hours by
        default).  This safeguard prevents Postfix processes from running
        indefinitely and consuming resources if a failure causes them to
        enter a stuck state.  XXX IMPROVE THIS SOMEHOW\@.

@@

The circumstances above account for all occasions identified thus far where
a \daemon{smtpd} suddenly stops logging.  In addition to removing an active
connection the last accepted mail may need to be discarded, as detailed in
\sectionref{timeouts-during-data-phase}; otherwise the queueid state table
is untouched.

These occurrences are handled by the three actions POSTFIX\_RELOAD,
SMTPD\_DIED, and SMTPD\_WATCHDOG\@.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
occurs only five times in 60,721,709 log lines, but for parser correctness
it must be dealt with.  The five occurrences have the same characteristics:
the \daemon{local} log line showing delivery to a local mailbox occurs
after the \daemon{qmgr} log line showing removal of the mail from the queue
because delivery is completed.  This causes problems: the data in the state
tables for the mail is not complete, so entering it into the database
fails; a new mail is created when the \daemon{local} line is parsed and
remains in the state tables; four warnings are issued per pair of out of
order log lines.

The COMMIT action examines the list of programs that have logged lines for
each mail, comparing the list against a table of known-good program
combinations.  If the mail's combination is found in the table the mail can
be entered in the database; if the combination is not found entry must be
postponed and the mail flagged for later entry.  The SAVE\_DATA action
checks for that flag; if the additional log lines have caused the mail to
reach a valid combination then entry in the database will proceed,
otherwise it must be postponed once more.

@@

XXX WOULD IT BE BETTER TO BEGIN THE EXPLANATIONS ON THE FOLLOWING LINE\@?

@@

    \item [\daemon{local}:] Local delivery of a bounce notification, or
        local delivery of a re-injected mail.

@@

    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Mail submitted
        locally on the server, \newline{} both local and remote delivery.

@@

    \item [\daemon{pickup}, \daemon{smtp}:] Mail submitted locally on the
        server, remote delivery only.

    \item [\daemon{smtp}:] Remote delivery of either a re-injected mail or
        a bounce notification.

    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
        client, then remotely delivered (typically relaying mail for
        clients on the local network to addresses outside the local
        network).

@@

This check applies to accepted mails only, not to rejected mails.  This
check is performed during the COMMIT action.

@@

    \item The timestamp of the last log line for the mail must be 12 hours
        or more earlier than the last log line parsed from the current log
        file.

    \item IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LINES\@?  There
        must be exactly two \daemon{smtpd} and one \daemon{cleanup} log
        entries for the mail, with no additional log entries.

@@

12 hours is a somewhat arbitrary time period, but it is far longer than
Postfix would delay delivery of a mail in the queue (unless Postfix is not
running for an extended period of time).  The state tables are scanned for
mails matching the characteristics above each time the end of a log file is
reached, and matching mails are deleted.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally
mails that are in the process of being delivered will be deleted.  This
results in the log lines from the delivery agent (\daemon{local},
\daemon{virtual} or \daemon{smtp}) appearing in the log file
\textit{after\/} the mail has been removed from the state tables and saved
in the database.  The DELETE action adds deleted mails to a cache, which is
checked by the SAVE\_DATA action, and the current log line discarded if the
following conditions are met:

@@

    \item The timestamp of the log line is within 5 minutes of the final
        timestamp of the mail.
